---
title: "Project 1 Report"
author: |
    | Analyst: Max McGrath
    | Investigator: Nichole Carlson
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
output: pdf_document
indent: true
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
library(kableExtra)
setwd("..")
source("Code/1_prepData.R")
source("Code/2_EDA.R")
source("Code/6_BA_AnalyzeChains.R")
```

\newpage

## Introduction

We received data of 3935 observations for 715 patients with HIV who began 
receiving highly active antiretroviral treatment (HAART). The data includes 
baseline values for each individual prior to starting treatment, then up 
to 8 years of longitudinal measurements for the patients as they received
HAART. Each observation includes laboratory and quality of life
measures, as well as demographic and other health information. Most relevant to
this analysis, each observation indicates whether patients used
illicit drugs via injection or opiates via any method of
administration, which will be referred to as "hard drug use" or just "drug use" 
throughout this analysis.

The goal of this analysis is to identify whether HAART deferentially impacts
patients who report using drugs prior to receiving treatment relative to those
who did not report using drugs at baseline.
The efficacy of HAART will be evaluated using four treatment outcomes: (1) viral
load, (2) CD4+ T cell counts, (3) aggregate physical quality of life score, and 
(4) aggregate mental quality of life score. We will test the hypothesis that
patients who report using illicit drugs prior to beginning treatment have worse
health outcomes after two years of treatment than those who did not report
using hard drugs at baseline.

## Methods

Patients were only included in this analysis if they had no missing values
for any of the four treatment outcomes, the indicator of whether they used hard 
drugs at baseline, or any of the baseline covariates described below. Values
that were deemed implausible were also removed. Viral load was also log base
10 transformed to account for the long-tailed distribution of those values.

Data was first summarized for relevant covariates and outcomes 
at both baseline and after
two years of treatment. Treatment outcomes were then analyzed using linear
models in both a frequentist context (multivariable least squares regression) 
and using Bayesian linear regression. For each framework four models per 
outcome were used (a total of sixteen models per framework), where differences 
in each outcome (year two values minus baseline) are used as the response for the 
respective models.

Hard drug use at baseline is the primary variable of interest, so each model
was defined to better understand the relationship between hard drug use and 
each respective outcome. First, models with only hard drug use and baseline
outcome values were estimated for each outcome. We will refer to those models
as univariable models. Next, confounding variables
were added to those models to control for factors that are likely correlated
with both drug use and the outcomes. Those confounding variables include
age, body mass index, smoking status, education, race/ethnicity. All confounding 
values were those measured at baseline, except adherence which used second year
values. These models will be refered to as multivariable models. Lastly, for each
outcome, the univariable and multivariable models were run without hard drug
use as a predictor to quantify the impact of removing hard drug use as a 
predictor. We will refer to these models as univariable without drug use and
multivariable without drug use, respectively.

For the Bayesian model, all priors were non-informative to reflect a lack of
*a priori* knowledge surrounding the associations between predictors and outcomes.
Specifically, coefficient priors were Gaussian with a mean of 0 and a
variance of 100,000 and error priors were gamma distributions with shape and
rate parameters both equal to 0.001. For each Bayesian model, two Monte Carlo Markov 
Chains were run for 25,000 iterations of Gibbs sampling to draw independent
values from each coefficient's posterior density to estimate those 
densities. The convergence and mixing of the chains was evaluated using 
trace plots, autocorrelation function plots, and by visually examining plots of
the posterior density estimates, all to ensure that the chains accurately
reflect the underlying parameters' distributions. After the chains were examined,
each respective pair was combined.

For the Bayesian linear models, parameter estimates, 95% highest posterior density
intervals (HPD), penalized deviance information criteria (DIC), and posterior 
probabilities of coefficients corresponding to a 10% or greater difference from
the sample mean outcome values were all calculated and evaluated to
evaluate the impact of hard drug use upon HAART. Those statistics'
frequentist counterparts, least-squares coefficient estimates, confidence
intervals, and t-test p-values were also used to evaluate the same relationship.
Concordance or lack thereof between the two paradigms as well as any limitations
to this analysis are also discussed.

## Results

Following the removal of incomplete observations, a total sample size of 463
patients remain, 36 who used hard drugs at baseline and 427 who did not. After
running the Markov Chains as described above and examining trace plots, 
autocorrelation function plots, and estimated posterior distribution plots, it
is clear that the chains are an accurate estimate of the posterior 
densities given the observed data and selected priors. Summaries of the
data used in this analysis are found in Table 1, and a missingness summary
for relevant baseline and year two values is found in Table 2.

Below we describe the results of the Bayesian modeling and contrast those results
with those of the frequentist modeling, with full results available in the
Appendix in Tables 3 and 4, respectively.

### Viral Load

For the univariate model, a coefficient estimate for drug use (the mean of the
posterior density) of 0.038 (95% HPD: -0.355, 0.429) 
was calculated. Since viral load was log base 10
transformed, we back-transform this estimate to arrive at the value 1.091, 
implying that drug use is, on average,
associated with 1.091 times higher difference in HIV copies in a mL of blood after
two years of HAART, relative to non-drug users. For the 
univariable model, the DIC was estimated to be 1438.458, while the same model
without drug use had a DIC of 1437.448. The posterior probability of the drug
use coefficient corresponding with a 10% change of the sample mean of viral 
load was calculated to be 0.828.

In both the univariate and multivariable models, HPDs included 0, the DIC
changed very little when drug use was removed from the model, and posterior
probabilities were insufficient to indicate significance, all indicating the
data does not support an association between drug use and change in viral load 
after two years of HAART. These results are further supported by the 
frequentist modeling.

### CD4+ T Cell Count

For the univariate model with CD4+ T cell count as the outcome, 
the drug use coefficient for drug use was estimated
to be -170.989 (95% HPD: -232.818, -112.506), suggesting that on average
drug users have -170.989 lower difference in CD4+ T cells than non-drug users 
after 2 years of
HAART. The DIC for the univariate model was 28.993 lower than the model without
drug use as a variable, indicating the inclusion of drug use notably improved
the fit of the model. The posterior probability for the CD4+ coefficient
indicating a 10% change in CD4+ counts was approximately 1.

Those results remained consistent for the multivariable model, both with and
without drug use as a predictor, indicating that the results were robust to the
inclusion of potential confounders. The results are also consistent with the
frequentist analysis.

### Mental Quality of Life Score

For the univariate model with mental quality of life score as the outcome,
the coefficient for drug use was estimated to be -0.136, but the 95% HPD
(-3.526, 3.250) is practically centered at 0. This, coupled with the only
slight decrease in DIC when including drug use as a coefficient, indicate that
the data does not support an association between drug use and differing 
changes in mental quality of life score between drug users and non-users after
two years of HAART.

This result is consistent with both the multivariable model and the frequentist
analysis.

### Physical Quality of Life Score

For the univariate model with physical quality of life score as the outcome,
the coefficient for drug use was estimated to be -4.074 (95% HPD: -6.691, -1.312)
with a posterior probability of drug use corresponding with an average 10% change
in physical quality of life score of 0.9988. The DIC with the inclusion of drug
use as a variable is slightly lower than that of the model without it. 

These results are somewhat consistent in the multivariable model, although the
HPD moves to include 0 (-5.433, 0.050) and the posterior probability of a 10%
change drops to 0.985, both indicating some level of confounding between the
additional variables and drug uses relationship with physical quality of life.

The results are consistent with those from the frequentist analysis,
with similar values for coefficient estimates, AIC, and confidence intervals
as their analogous Bayesian statistics. In both cases, after controlling for
potential confounders, there remains weak evidence that drug use
and change in physical quality of life score after 2 years of HAART are 
negatively associated.

## Conclusions

From the above results, we see strong evidence that differences in CD4+ cell
counts after two years of HAART are smaller in individuals that use hard drugs,
implying there may be a negative effect of hard drug use on HAART. We also see
weak evidence that drug users see reduced changes in physical quality of life
score after 2 years of HAART relative to non-drug users. We do not, however,
note any significant difference in viral load or mental quality of life score
between individuals who use hard drugs and those who do not. As such, we do
conclude that drug use is associated with worse health outcomes after two years
of HAART.

### Limitations and Discussion

The greatest limitation of this study is the small sample of drug users relative
to non-users. When considering this difference it must be noted that if it is 
the case that hard drug use negatively impacts
the efficacy of HAART, there are several factors which would likely bias the
above results. First, the use of hard drugs was self-reported by individuals
being treated. Given the gravity of the social stigma surrounding drug usage
(particularly the drugs being studied), it is foreseeable that some users would
be disinclined to report their drug use, leading to the presence of drug users
in the non-drug using study arm, thereby diminishing the measured effect.
Further, it would be foreseeable that dropout rate among drug users would be 
higher than among non-users due to both diminished health (and death) and to
socioeconomic factors, which would both bias our results, as those with the
worst health out comes may drop out, and reduce the sample size of drug users
in the study.

Given these factors, the above findings should be 
considered with greater leniency, where even weak effects should be considered
indicative that a meaningful correlation may exist. As such, our results are 
indicative of a meaningful correlation between drug use and worse health outcomes
are 2 years of HAART. Further research that aims to expand the
quantity of drug using patients should be pursued. This could be accomplished
through more active recruiting and provision of HAART treatment to drug using
individuals and through refinement of data collection methods to foster
a more welcoming environment in which individuals may feel more inclined to
self-report their drug use.

\newpage

## Reproducibility

The code used to generate this analysis is available on GitHub 
at https://github.com/BIOS6624-UCD/bios6624-MaxMcGrath/tree/main/Project1
. The `Background` folder contains information pertinent to understanding the
analysis but unnecessary for reproducing it. The `Code` folder contains six
files: `1_prepData.R`, `2_EDA.R`, `3_FrequentistAnalysis.R`, 
`4_BA_DrawChains.R`, `5_BA_AssessChains.R` and `6_BA_AnalyzeChains.R`. 
These R scripts are dependent upon a data file 
`Data/hiv_6624.csv` which is not available on GitHub, but may be 
requested
by emailing `max.mcgrath@ucdenver.edu`. To run the complete analysis, each
script should be run in the order of the number preceding its filename. The
last directory, `Report`, contains the RMarkdown file `report.Rmd` which may
be used to generate this report (note that it also depends on the aforementioned
data and scripts).

The complete details of the `R` version, package versions, and
machine details for the instance which generated this report are provided below.

```{r, echo = FALSE}
sessionInfo()
```


\newpage

```{r, echo = FALSE}
# Create Table 1
kable(table1(~ LOG_VLOAD_DIFF + LEU3N_DIFF + MENT_DIFF + PHYS_DIFF +
           AGE_0 + BMI_0 + RACE_0 + EDUC_0 + SMOKE_0 + ADH_2 +
           LOG_VLOAD_0 + LEU3N_0 + MENT_0 + PHYS_0 | DRUGS_0, 
       data = tableOneData, render.categorical="FREQ (PCTnoNA%)",
       caption = "Data Summary"),
       caption = "Data Summary") %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```

```{r, echo = FALSE}
kable(missingSummary, booktabs = TRUE, caption = "Missingess Summary",
      digits = 2, col.names = c("Variable", "N Missing", "% Missing", "N Missing", "% Missing")) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>%
    add_header_above(c(" " = 1, "Baseline" = 2, "Year Two" = 2))
```

```{r, echo = FALSE}
kable(fullBayesianSummary[, 2:7], booktabs=TRUE, 
      caption = "Bayesian Analysis Summary", 
      digits = 4, col.names = c("Model", "Penalized DIC", "Estimate", 
                                "HPD Lower", "HPD Upper", 
                                "Posterior Probability")) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>%
    add_header_above(c(" " = 2, "Drug Use Coefficient" = 4)) %>%
    pack_rows("Standardized Viral Load (log10 copies/ml)", 1, 4) %>%
    pack_rows("CD4+ cells", 5, 8) %>%
    pack_rows("SF36 MCS score", 9, 12) %>%
    pack_rows("SF36 PCS score", 13, 16)
```


```{r, echo = FALSE}
kable(fullFreqSummary[, 2:7], booktabs=TRUE, 
      caption = "Frequentist Analysis Summary",
      digits = 4, col.names = c("Model", "AIC", "Estimate", "Lower CI", 
                                "Upper CI", "p-value")) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>% 
    add_header_above(c(" " = 2, "Drug Use Coefficient" = 4)) %>%
    pack_rows("Standardized Viral Load (log10 copies/ml)", 1, 4) %>%
    pack_rows("CD4+ cells", 5, 8) %>%
    pack_rows("SF36 MCS score", 9, 12) %>%
    pack_rows("SF36 PCS score", 13, 16)
```