---
title: "Interim Data Analysis Plan"
author: |
    | Analyst: Max McGrath
    | Investigator: Nichole Carlson
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
output: pdf_document
indent: true
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
library(kableExtra)
```

\newpage

## Introduction

This report will assist in preparing a grant application by developing
an analysis plan, sample size estimation, and a budget with justification 
for a study
by Dr. Brianne Bettcher seeking to understand the relationship between 
inflammation, Alzheimer's disease (AD), and cognitive decline. This report
will contribute to Aims 1 and 2 of the grant.

## Aim 1 Analysis

For the first aim, we will use linear regression models to analyze the associations, 
with changes in memory and cortical thickness in one year as the 
outcomes and cytokines and chemokines as predictors. We will also
adjust for age and sex as potential confounders, and potentially other potential
confounders if the need arises. We will use coefficient estimates and t-tests to
evaluate each variable as a predictor, along with some kind of multiple testing
correction to account for the number of t-tests performed.

## Aim 2 Analysis

For both parts of the second aim, we will use linear models to 
analyze the relevant predictive power and associations. For the first part, we will
use amyloid deposition and cortical thickness as the outcomes with inflammatory
markers as predictors. For part the second, we will use change in memory and cortical
thickness in one year as outcomes and amyloid deposition, inflammatory markers,
and cytokines and chemokines as variables of interest. In both cases, we will
adjust for sex and age as potential confounders, and use coefficient estimates
and t-tests to evaluate significance and associations. As with Aim 1, we will
use multiple testing correction to account for the number of t-tests performed. 

## Sample Size Estimation

We will estimate the necessary sample size to achieve 80% power based on 
preliminary estimates of correlations between 0.25 and 0.4 between
inflammatory markers and cognition and a chosen significance level of 0.05. 
Several potential standard deviation estimates for aCMI change per year have
also been provided. Those estimates will be evaluated, and at least one will
be used in the sample size estimation. The estimate will also account for
multiple comparisons. Both aims use the same sample, so the estimate will be
based on the aim requiring a larger sample size and will therefore be sufficient
for both aims. The estimate will then be used to either support Dr. Bettcher's
proposed sample size or to 
suggest modifications to that sample size.

## Budget

Lastly, a budget with justification will be developed for the project using a 
maximum budget of $500,000 per year for four years. Tables 1 and 2 will be 
filled in
based on estimated effort levels by all listed personnel and their provided
salary estimates (note that the tables will eventually be combined into a single
table).

\newpage

```{r, include = FALSE}
position <- c("PhD Statistician", "MS Statistician", "Data Manager",
              "Research Assitant", "Totals")
baseSalary <- c(115000, 80000, 65000, 43000, NA)
salaryWithBenefits <- c(115000*1.28, 80000*1.28, 65000*1.28, 43000, NA)
yearOneEffort <- c("-", "-", "-", "-", NA)
yearOneCost <- c("-", "-", "-", "-", "-")
yearTwoEffort <- c("-", "-", "-", "-", NA)
yearTwoCost <- c("-", "-", "-", "-", "-")
yearThreeEffort <- c("-", "-", "-", "-", NA)
yearThreeCost <- c("-", "-", "-", "-", "-")
yearFourEffort <- c("-", "-", "-", "-", NA)
yearFourCost <- c("-", "-", "-", "-", "-")
totals <- c("-", "-", "-", "-", "-")
budgetDF1 <- data.frame(position, baseSalary, salaryWithBenefits, yearOneEffort,
                        yearOneCost, yearTwoEffort, yearTwoCost)
budgetDF2 <- data.frame(position, yearThreeEffort, yearThreeCost, 
                        yearFourEffort, yearFourCost, totals)
```

```{r, echo = FALSE}
kable(budgetDF1, booktabs = TRUE, caption = "Budget Justification",
      col.names = c("Position", "Base Salary", "Salary w/ Benefits", 
                    "Y1 Effort", "Y1 Cost", "Y2 Effort", "Y2 Cost")) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```
```{r, echo = FALSE}
kable(budgetDF2, booktabs = TRUE, caption = "Budget Justification cont'd",
      col.names = c("Position", "Y3 Effort", "Y3 Cost", "Y4 Effort", "Y4 Cost",
                    "Totals")) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```

