---
title: "Project 2 Report"
author: |
    | Analyst: Max McGrath
    | Investigator: Nichole Carlson
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
output: pdf_document
indent: true
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
library(kableExtra)
library(xtable)
library(stringr)
```

\newpage

## Introduction

This report will assist in preparing a grant application by developing
an analysis plan, sample size estimation, and a budget with justification 
for a study
by Dr. Brianne Bettcher seeking to understand the relationship between 
inflammation, Alzheimer's disease (AD), and cognitive decline. This report
will contribute to Aims 1 and 2 of the grant.

## Aim 1 Analysis

For the first aim, we will use linear regression models to analyze the associations, 
with changes in memory and cortical thickness in one year as the 
outcomes and baseline cytokines and chemokines as predictors for the first part 
and changes
in cytokines and chemokines in one year as predictors for the second part. In
both models we will control for age and sex as confounders.

We have also been given a large number of potentially confounding factors such
as inflammation, cardiovascular risk, health history, and APOE 
genotype. Given the number of potential confounders coupled with the foreseeable
collinearities between confounders, we will use elastic net regressions with
potential confounders as predictors and each aforementioned outcome as
responses to identify a final set of confounders which will be included in the
above linear regression models.

We will then use coefficient estimates and t-tests to
evaluate each variable as a predictor, using a Bonferroni corrected p-value to
account for the number of t-tests performed.

## Aim 2 Analysis

For both parts of the second aim, we will use linear models to 
analyze the relevant predictive power and associations. For the first part, we will
use amyloid deposition and cortical thickness as the outcomes with inflammatory
markers as predictors. For the second part, we will use changes in memory and cortical
thickness in one year as outcomes and amyloid deposition, inflammatory markers,
and cytokines and chemokines as variables of interest. We will also include an
interaction term between amyloid deposition and inflammatory markers to test the
hypothesis that inflammatory markers differentially impact the outcomes depending
on amyloid deposition, where amyloid deposition is treated as a dichotomous variable
(high/low levels).

As with Aim 1, we will control for age and sex and use elastic net regression to
expand the set of
confounders to include in the above linear models. We will then use coefficient 
estimates and t-tests to
evaluate each variable as a predictor, using a Bonferroni corrected p-value to
account for the number of t-tests performed.

## Sample Size Estimation

Using change in recall on a episodic memory list-learning test (CVLT) as the 
primary outcome of interest, to achieve 80% power with a Bonferroni corrected 
p-value of 0.0083 (based on a desired type 1 error rate of 0.05 and tests for 
6 cytokines and chemokines of interest), assuming
a difference in correlation of 0.4, a sample size of 276 (138 with high amyloid deposition
and 138 with low deposition) is estimated to be necessary. Then, accounting
for 10% one-year dropout, a total sample size of 304 is estimated to be sufficient
to achieve 80% power. 

This estimate was powered for the second part of Aim 2, where the linear model
described above was simplified to a two independent correlation z-test. This
sample size estimate leads to an estimated power of just under 1 for Aim 1 based
on a simplification of the linear models to a correlation z-test. As such, a
sample size of 304 is estimated to be sufficient for both aims with allowance
for 10% attrition. This is notably higher than Dr. Bettcher's initial estimate of 194
subjects, so more enrollment is necessary. Note that this is an anti-conservative
estimate, as we plan to add confounders to each model which may reduce power.

## Budget

Tables 1 and 2 provide a budget estimate and budget justification for the statisticians
and data managers involved in the 5-year project. The budget will fund one PhD
statistician and one MS statistician who will be involved with study design,
implementation, data collection, analysis, and regular meeting attendance
throughout the duration of the grant. The PhD statistician will primarily be
responsible for assembling the rest of the statistical personnel and supervising
and guiding their work, while the MS statistician will be primarily responsible
for analysis and manuscripts.
The budget will also fund a data manager who will primarily be tasked with the creation
of a REDCap data set, data management, and data documentation
in the first two years of the grant,
with some data set maintenance in the latter years. Lastly, the budget will
fund a student researcher who will assist with data management and analysis.

Estimated personnel salaries are provided in Table 1, and yearly personnel effort
estimates along with associated costs are provided in Tables 1 and 2. 
The total statistical budget for the grant is $92,670 for the first year,
$70,770 for the second,
$52,350 for the third,
$88,700 for the fourth,
and $92,860 for the fifth year. The total budget of the 5-year duration of the
grant is $397,350.

\newpage

```{r, include = FALSE}
position <- c("PhD Statistician", "MS Statistician", "Data Manager",
              "Research Assistant", "Totals")
baseSalary <- c(115000, 80000, 65000, 43000, NA)
salaryWithBenefits <- c(115000*1.28, 80000*1.28, 65000*1.28, 43000, NA)
yearOneEffort <- c(.1, .25, .5, .25, NA)
yearOneCost <- c(salaryWithBenefits[1]*yearOneEffort[1], 
                 salaryWithBenefits[2]*yearOneEffort[2], 
                 salaryWithBenefits[3]*yearOneEffort[3], 
                 salaryWithBenefits[4]*yearOneEffort[4], 
                 salaryWithBenefits[1]*yearOneEffort[1] + 
                 salaryWithBenefits[2]*yearOneEffort[2] +  
                 salaryWithBenefits[3]*yearOneEffort[3] +  
                 salaryWithBenefits[4]*yearOneEffort[4])
yearTwoEffort <- c(.05, .15, .5, .15, NA)
yearTwoCost <- c(salaryWithBenefits[1]*yearTwoEffort[1], 
                 salaryWithBenefits[2]*yearTwoEffort[2], 
                 salaryWithBenefits[3]*yearTwoEffort[3], 
                 salaryWithBenefits[4]*yearTwoEffort[4], 
                 salaryWithBenefits[1]*yearTwoEffort[1] +
                 salaryWithBenefits[2]*yearTwoEffort[2] + 
                 salaryWithBenefits[3]*yearTwoEffort[3] + 
                 salaryWithBenefits[4]*yearTwoEffort[4])
yearThreeEffort <- c(.15, .15, .05, .25, NA)
yearThreeCost <- c(salaryWithBenefits[1]*yearThreeEffort[1], 
                 salaryWithBenefits[2]*yearThreeEffort[2], 
                 salaryWithBenefits[3]*yearThreeEffort[3], 
                 salaryWithBenefits[4]*yearThreeEffort[4], 
                 salaryWithBenefits[1]*yearThreeEffort[1] + 
                 salaryWithBenefits[2]*yearThreeEffort[2] + 
                 salaryWithBenefits[3]*yearThreeEffort[3] + 
                 salaryWithBenefits[4]*yearThreeEffort[4])
yearFourEffort <- c(.15, .4, .05, .5, NA)
yearFourCost <- c(salaryWithBenefits[1]*yearFourEffort[1], 
                 salaryWithBenefits[2]*yearFourEffort[2], 
                 salaryWithBenefits[3]*yearFourEffort[3], 
                 salaryWithBenefits[4]*yearFourEffort[4], 
                 salaryWithBenefits[1]*yearFourEffort[1] + 
                 salaryWithBenefits[2]*yearFourEffort[2] + 
                 salaryWithBenefits[3]*yearFourEffort[3] + 
                 salaryWithBenefits[4]*yearFourEffort[4])
yearFiveEffort <- c(.15, .4, .1, .5, NA)
yearFiveCost <- c(salaryWithBenefits[1]*yearFiveEffort[1], 
                 salaryWithBenefits[2]*yearFiveEffort[2], 
                 salaryWithBenefits[3]*yearFiveEffort[3], 
                 salaryWithBenefits[4]*yearFiveEffort[4], 
                 salaryWithBenefits[1]*yearFiveEffort[1] + 
                 salaryWithBenefits[2]*yearFiveEffort[2] + 
                 salaryWithBenefits[3]*yearFiveEffort[3] + 
                 salaryWithBenefits[4]*yearFiveEffort[4])
totals <- c(yearOneCost[1] + yearTwoCost[1] + yearThreeCost[1] + yearFourCost[1] + yearFiveCost[1], 
            yearOneCost[2] + yearTwoCost[2] + yearThreeCost[2] + yearFourCost[2] + yearFiveCost[2], 
            yearOneCost[3] + yearTwoCost[3] + yearThreeCost[3] + yearFourCost[3] + yearFiveCost[3],
            yearOneCost[4] + yearTwoCost[4] + yearThreeCost[4] + yearFourCost[4] + yearFiveCost[4],
            yearOneCost[1] + yearTwoCost[1] + yearThreeCost[1] + yearFourCost[1] + yearFiveCost[1] +
            yearOneCost[2] + yearTwoCost[2] + yearThreeCost[2] + yearFourCost[2] + yearFiveCost[2] + 
            yearOneCost[3] + yearTwoCost[3] + yearThreeCost[3] + yearFourCost[3] + yearFiveCost[3] +
            yearOneCost[4] + yearTwoCost[4] + yearThreeCost[4] + yearFourCost[4] + yearFiveCost[4])

# Format percentages and dollar amounts
yearOneCost <- paste0("$", format(yearOneCost, big.mark=",", flag="-"))
yearTwoCost <- paste0("$", format(yearTwoCost, big.mark=",", flag="-"))
yearThreeCost <- paste0("$", format(yearThreeCost, big.mark=",", flag="-"))
yearFourCost <- paste0("$", format(yearFourCost, big.mark=",", flag="-"))
yearFiveCost <- paste0("$", format(yearFiveCost, big.mark=",", flag="-"))
baseSalary <- paste0("$", format(baseSalary, big.mark=",", flag="-"))
salaryWithBenefits <- paste0("$", format(salaryWithBenefits, big.mark=",", flag="-"))
totals <- paste0("$", format(totals, big.mark=",", flag="-"))

yearOneCost <- str_replace(yearOneCost, " ", "")
yearTwoCost <- str_replace(yearTwoCost, " ", "")
yearThreeCost <- str_replace(yearThreeCost, " ", "")
yearFourCost <- str_replace(yearFourCost, " ", "")
yearFiveCost <- str_replace(yearFiveCost, " ", "")
baseSalary <- str_replace(baseSalary, " ", "")
salaryWithBenefits <- str_replace(salaryWithBenefits, " ", "")
totals <- str_replace(totals, " ", "")

yearOneEffort <- paste0(yearOneEffort*100, "%")
yearTwoEffort <- paste0(yearTwoEffort*100, "%")
yearThreeEffort <- paste0(yearThreeEffort*100, "%")
yearFourEffort <- paste0(yearFourEffort*100, "%")
yearFiveEffort <- paste0(yearFiveEffort*100, "%")

# Reset NAs
baseSalary[5] <- NA
salaryWithBenefits[5] <- NA
yearOneEffort[5] <- NA
yearTwoEffort[5] <- NA
yearThreeEffort[5] <- NA
yearFourEffort[5] <- NA
yearFiveEffort[5] <- NA


# Create data frame of budget data
budgetDF1 <- data.frame(position, baseSalary, salaryWithBenefits, yearOneEffort,
                        yearOneCost, yearTwoEffort, yearTwoCost)
budgetDF2 <- data.frame(position, yearThreeEffort, yearThreeCost, 
                        yearFourEffort, yearFourCost, yearFiveEffort,
                        yearFiveCost, totals)
```

```{r, echo = FALSE}
kable(budgetDF1, booktabs = TRUE, caption = "Budget Justification",
      col.names = c("Position", "Base Salary", "Salary w/ Benefits", 
                    "Y1 Effort", "Y1 Cost", "Y2 Effort", "Y2 Cost")) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```
```{r, echo = FALSE}
kable(budgetDF2, booktabs = TRUE, caption = "Budget Justification cont'd",
      col.names = c("Position", "Y3 Effort", "Y3 Cost", "Y4 Effort", "Y4 Cost",
                    "Y5 Effort", "Y5 Cost", "Totals")) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```