---
title: "Project 3 Report"
author: |
    | Analyst: Max McGrath
    | Investigator: Nichole Carlson
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
output: pdf_document
indent: true
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(gridExtra)
options(knitr.kable.NA = '')
setwd("..")
source("Code/2_EDA.R")
source("Code/3_KM.R")
source("Code/5_ProbabilitiesM.R")
source("Code/6_ProbabilitiesF.R")
source("Code/7_TimeAnalysis.R")
```

\newpage

## Introduction

We received data of 11,627 observations from 4,434 individuals with up to 
three observations per individual tracking subjects'
cardiovascular outcomes for up to 24 years. Most relevant to this study, each
observation records whether the subject has had a stroke since the last observation 
and, if so, the timing of that stroke. Each observation also measures four continuous 
cardiovascular risk factors: (1) age, (2) systolic blood pressure (SBP), (3) cholesterol 
levels, and (4) BMI, and four dichotomous variables: (5) smoking status, (6)
diabetic status, (7) history of cardiovascular disease (CVD), and (8) use of
anti-hypertension medication.

The aim of this study will be to use this data to identify statistically 
significant risk factors that are associated with greater risk of stroke 10 years
after the start of observation by testing the hypotheses that each of the eight
previously listed risk factors are associated with 10-year risk of stroke.
The second aim is to determine 10-year probabilities of 
strokes for different risk profiles using the significant risk factors. 
The last aim of this analysis is to
identify whether their is meaningful change in risk factors over the first 10
years of the study to determine whether a longitudinal analysis approach is
warranted. This study is stratified by sex, meaning
males and females are treated as two separate populations with potentially
differing risk factors. 

## Methods

First, subjects who have had a stroke prior to the start of observation were 
removed from the data set. Next, for each subject, a record of whether they
had a stroke in the first 10 years of observation was created along with the
timing of that stroke (which we will refer to as events and event times, 
respectively). If they did not 
have an event in those years, they were classified as right censored. Based on 
the findings of Wolf et al. (1990) as well as data availability, age, systolic
blood pressure (mmHg) (SBP), use of anti-hypertensive medication, diabetes 
mellitus, cigarette smoking, past cardiovascular disease (CVD), BMI, and serum 
total cholesterol (mg/dL) were included in the analysis
as potential risk factors (all measured at baseline). Age, SBP, cholesterol, and
BMI will be treated as continuous variables with the rest treated as 
dichotomous. Only observations with no missing values for any of the potential
risk factors were used.

The first part of the analysis provides descriptive statistics for
potential risk factors, stratified by sex and whether the 
subject had an event or was censored. Next, estimated Kaplan-Meier survival
curves were fit to crudely estimate event probability over time. For 
continuous risk factors, the K-M curves were fit according to the risk factor
quantiles. Then, to 
fully quantify each risk factor's relationship with stroke 
occurrence, a Cox proportional hazards (CPH)
regression was fit for each sex with the number of days after beginning 
observation that a stroke occurred or the last observation time in days as the 
outcome and the potential risk factors described above as variables of interest.
The risk factors were then narrowed using backwards selection with statistical
significance (alpha $<0.05$) as the criteria for whether each factor was kept in 
the model. 
After predictor selection, a final CPH model was fit for each sex.
Coefficient estimates from these models were then used to quantify
the relationship between risk factors and stroke occurrence, with
Wald tests quantifying these relationships' statistical significance. 10-year
stroke probabilities were then calculated for a predefined set of risk profiles.

Lastly, we provide summary statistics of the changes in potential risk 
factors from the first observation to the second and discuss the potential
impact of using a more complex, time-varying covariate analysis rather than
simply evaluating relationships between baseline values and stroke occurrence.
As a part of this discussion, the proportional hazards assumption for Cox
proportional hazards models was evaluated in the context of this study to
identify whether the relationship between covariates and stroke risk varies
with time thereby altering the interpretation of this study's results.

## Results

After preparing the data, there were 2378 subjects with sex listed as female
including 57 who had an event and 2321 who are censored, and there were 1897 
subjects with sex listed as male with 48 who had an event and 1849 who are 
censored.

From Table 1, the majority of individuals observed did not have
strokes within 10 years of observation (2.5% of males and 2.4% of 
females had strokes). Among
those who did have strokes, generally they are older, have higher SBP, are
more likely to use anti-hypertensive medication, are more likely to have
diabetes, and are more likely to have a history of CVD. From the Kaplan-Meier 
survival curves shown in Figures 1 and 2, age, SBP, and
all dichotomous measures appear to be associated with greater incidence of
strokes in males, and all continuous measures and all dichotomous measures except
current smoking status appear to be associated with greater incidence of strokes
in females. Both the descriptive statistics and Kaplan-Meier curves lend 
credence to the selection of the initial set of potential risk factors to 
consider.

Following backwards model selection, age, SBP, diabetes, and smoking status
remained as significant predictors of stroke incidence in males, while only
age and SBP remained as significant predictors of stroke incidence in females.
For males, having diabetes at baseline was associated with an estimated hazard 
of stroke within 10 years of observation of 4.79 (95% CI: (2.1, 10.99)) times 
that of those who did not have diabetes at baseline holding age, 
BPD, and smoking status constant (p-value $< 0.001$), and being a smoker at 
baseline was associated with an estimated hazard of having a stroke of 1.92 
(95% CI: (1.03, 3.56))
times that of those who were not smokers at baseline, holding age, SBP, and
diabetic status constant (p-value $= 0.039$). For both males and females,
the estimated hazard of stroke was 1.07 times greater for each additional year
of age (95% CI's: (1.03, 1.11) and (1.03, 1.12), respectively), and the 
estimated hazard of stroke was 1.03 times greater for each additional mmHg of
blood pressure (95% CI's: (1.02, 1.05) and (1.02, 1.03), respectively) with
p-values $\leq.001$ for both age and SBP in each sex's respective models.

Table 3 provides the probabilities of stroke within 10 years for a variety 
of risk profiles as estimated by the Cox proportional models. 
The average probability of stroke
within ten years is 2% at 55 and 15% at 85 for males and 2% at 55 and 16% at 85 
for females. For females, the probability of stroke at each age remains constant
for each risk profile, as none of the reported risk profiles are significantly 
associated with stroke probability. For males, being a smoker and having 
diabetes both  increase the probability of having a stroke within 10 years for 
all ages, with the conjunction of the two carrying the highest probability of 
stroke within 10 years of any risk profile at all ages (13% at 55, 61% at 85).

For the longitudinal comparison between the first and second period, Table 4
provides descriptive statistics of the differences between the two periods. 
Only 1594 males and 2001 subjects that were present for period one (after
removing those who had previously had a stroke) were also present for period two.
For the continuous risk factors, BMI slightly increased for those who were
censored and decreased for those who had strokes, while cholesterol and SBP
increased for both censored and those who had a stroke. For dichotomous factors,
more individuals stopped smoking than started smoking, and similarly more 
individuals started taking anti-hypertension medication than stopped taking it.
There was also a small percentage of individuals who were recognized as diabetic
between the first and second period or who had a history of CVD in the second
period but not in the first. The proportional hazards assumption of the CPH
models were also evaluated using Schoenfeld tests, and for both the male and 
female models there was no evidence of violation of the proportional hazards
assumption (p-values of 0.35 and 0.95, respectively).

## Conclusions

From the CPH models, only age and systolic blood pressure were significantly
associated with increased hazard of stroke for females, and age, systolic blood
pressure, smoking status, and diabetic status were significantly associated with
increased hazard of stroke in males. From the 10-year stroke probabilities
calculated using these models, the probability of stroke increases with age
for all risk profiles. For females, the probability of stroke is constant across
all risk profiles for a fixed age, and for males those who are current smokers,
those who are diabetic, or those who are both are associated with increased 
10-year probabilities of stroke, the latter the most of any risk profile. 

From the comparison of the first and second period, the across-the-board 
increase in SBP coupled with SBP's significant association with stroke hazard is
indicative that a longitudinal approach may be preferable for fully assessing
stroke hazard, although the proportional hazards assumption for this analysis 
was met indicating a non-longitudinal approach may have been sufficient. 
A longitudinal analysis would also account for observed changes in
smoking habits, anti-hypertension medication use, and diabetic status.

## Limitations

The range of ages for which risk probabilities (55 - 85 years) were calculated 
was based upon the methods of Wolf (1990), but it may be problematic to extend
the inferences of this study to the higher ages in that range. Specifically, at
baseline the largest ages present in the study were 69 for male and 70 for
female subjects, with averages of about 50 years for both sexes. Coupled with
the low incidence of stroke among the studied populations, the studied data
may not provide helpful insight into higher age ranges. 

Further, several of
the risk factors that were identified as significant by Wolf et al. were not
identified as significant in this study, and no factors which Wolf et al. did
not consider were identified as significant. As such, this study is likely 
under powered and would be improved by increasing the sample size, particularly
to capture a greater number of individuals who have strokes in the period of
observation.

\newpage

## Reproducibility

The code used to generate this analysis is available on GitHub 
at https://github.com/BIOS6624-UCD/bios6624-MaxMcGrath/tree/main/Project3
. The `Background` folder contains information pertinent to understanding the
analysis but unnecessary for reproducing it. The `Code` folder contains six
files: `1_ProcessData.R`, `2_EDA.R`, `3_KM.R`, 
`4_ModelSelection.R`, `5_ProbabilitiesM.R`, `6_ProbabilitiesF.R`, and 
`7_TimeAnalysis.R`.
These R scripts are dependent upon a data file 
`Data/frmgham2.csv` which is not available on GitHub, but may be 
requested
by emailing `max.mcgrath@ucdenver.edu`. To run the complete analysis, each
script should be run in the order of the number prefixing its filename. The
last directory, `Report`, contains the RMarkdown file `report.Rmd` which may
be used to generate this report (note that it also depends on the aforementioned
data and scripts).

The complete details of the `R` version, package versions, and
machine details for the instance which generated this report are provided below.

```{r, echo = FALSE}
sessionInfo()
```

## Works Cited

Wolf, P.A., D'Agostino, R.B., Belanger, A.J., and Kannel, W.B. (1991) Probability
of Stroke: A Risk Profile From the Framingham Study. *Stroke, 22* (3), 312-318

\newpage

## Appendix 

```{r, echo = FALSE}
t1kable(tableOne, booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```

\newpage

```{r, echo=FALSE, fig.cap="Kaplan-Meier Survival Curves for Male Sex", fig.height=6.7}
arrange_ggsurvplots(kmPlotsM[2:9],
                    print = TRUE,
                    ncol = 2,
                    nrow = 4,
                    surv.plot.height = NULL,
                    risk.table.height = NULL,
                    ncensor.plot.height = NULL,
                    common.legend = FALSE)
```

\newpage

```{r, echo=FALSE, fig.cap="Kaplan-Meier Survival Curves for Female Sex", fig.height=6.7}
arrange_ggsurvplots(kmPlotsF[2:9],
                    print = TRUE,
                    ncol = 2,
                    nrow = 4,
                    surv.plot.height = NULL,
                    risk.table.height = NULL,
                    ncensor.plot.height = NULL,
                    common.legend = FALSE)
```

\newpage

```{r, echo=FALSE}
kable(cphSummaryDF, booktabs = TRUE,
      caption = "Model coefficients (stratified by sex)",
      col.names = c("Variable", "Hazard Ratio", "95% CI", "p-value",
                    "Hazard Ratio", "95% CI", "p-value"),
      row.names = FALSE) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>% 
    add_header_above(c(" " = 1, "Male" = 3, "Female" = 3))
```


```{r, echo=FALSE}
kable(survProbDF, booktabs = TRUE,
      caption = paste0("Probability of Stroke at 10 Years stratified by ", 
                       "baseline age in years and sex"),
      col.names = c("55", "60", "65", "70", "75", "80", "85",
                    "55", "60", "65", "70", "75", "80", "85"),
      row.names = TRUE) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>%
    column_spec(c(1,8), border_right = TRUE) %>%
    add_header_above(c(" " = 1, "Male" = 7, "Female" = 7))
```


```{r, echo=FALSE}
t1kable(diffTable, booktabs = TRUE) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
```

